A Type System for Safe and Scalable Parallelism using Logical Regions

### 2012-07-06 doc/popl2013/langdef.tex
intro
cleanup

\pagebreak

\section{Legion Core Language}
\label{sec:legioncore}

\subsection{Language Definition}
\label{subsec:langdef}

Figure~\ref{fig:langdef} defines the core subset of Legion application language.
Legion includes basic boolean, integer and tuple types.  Pointers are included as well, but in
addition to specifying the type of value it points to, a pointer's type also constrains where in
memory it can point (i.e. only into regions listed in the type).  A special type exists for
{\em colorings}, which are used to specify how a region should be partitioned into subregions.
Functions in Legion are named, accept one or more arguments and return a value of some type.
A function also declares what region access privileges and coherence it requires.  Function types
are universally quantified over all logical region bindings, with all argument and result types,
as well as the privileges and coherence requirements being expressed relative to the quantified
region names.
The final additional value type that Legion uses is a region relation.  Although logical regions
cannot be stored directly in variables, they can be captured as part of a region relation 
instance, which stores a value and one or more logical regions that satisfy a particular set of
constraints.  An instance of a region relation can be read out of the heap and unpacked, giving
new local names to the logical regions contained in the instance and tracking the constraints
between those new logical regions.

\subsection{Legion Core Type System}

The Legion Core language is explicitly typed, using judgments of the following form:

\begin{center}
$\typeenv{e}{T}$
\end{center}

where:
\begin{itemize}
\item $\Gamma$ is the type environment, mapping the names of local variables and global functions
to their types,
\item $\Phi$ is the region access privileges under which the expression will be evaluated, and
\item $\Omega$ is the set of constraints that must hold between regions
\end{itemize}

As regions are hierarchical, privileges and/or constraints on a region will often imply similar
privileges or constraints on subregions.  Figure~\ref{fig:closure} defines transitive closure
operations for $\Phi^*$ and $\Omega^*$ that will be used in some of the type checking rules.

The full set of type checking rules are in Figure~\ref{fig:types}.  Many of the rules are 
straightforward, but several deserve more explanation.

The Legion core language has no subtyping, even for pointers into regions that are part of a 
known region hierarchy.  The region ``scope'' of a pointer can only be changed via explicit
{\em upregion} and {\em downregion} operations, which check that the ``from'' and ``to'' regions
are in the right relationship.  It would have been possible to safely infer {\em upregion} calls,
but {\em downregion} calls must be accompanied by a run-time check that the pointer really does
point into the subregion(s), so {\em upregion} was made explicit as well for simplicity.

@@

Due to space constraints, we limit the discussion to the general strategy of the proof.  The full
proof can be found here\cite{xyz}.  If $v$ and $E$ are the result of a permissible evaluation of
$e$, then a proof exists of:
$$\opsenv{e}{v}{E}$$
The soundness of the result is shown by induction over the structure of this proof.  For each of 
the 23 different forms a Legion expression can have, we assume the soundness of the subexpressions'
evaluations (provided we can show the consistency of the mappings and store used by that
subexpression) and prove that the result of the expression's evaluation is sound.  Although there
are a multitude of cases to be proven, many of them are similar, and benefit from
the use of the following lemmas:

@@

The proofs of these lemmas can also be found here\cite{xyz}.  The proof of the theorem is organized
into six pieces, one for each of the three consistency properties needed for subexpressions and one
for each of the three soundness properties we wish to show.

>>>>

\section{Legion}
\label{sec:legioncore}

Figure~\ref{fig:langdef} defines the core subset of Legion.  Types
include booleans, integers, tuples, and pointers annotated with a list
of regions into which they may point.  There is a special type for
{\em colorings}, which are used to specify partitions of regions into
subregions.  Functions/tasks in Legion are named, accept one or more
arguments and return a value of some type.  A function also declares
what region access privileges and coherence it requires.  Function
types are universally quantified over all logical region bindings,
with all argument and result types, as well as the privileges and
coherence requirements being expressed relative to the quantified
region names.  The final additional value type that Legion uses is a
region relation, which stores a value and one or more logical regions
satisfying associated region constraints.  An instance of a region
relation can be read out of the heap and unpacked, giving new local
names to the logical regions contained in the instance.
% and tracking
%the constraints between those new logical regions.

\subsection{Core Type System}

Core Legion is explicitly typed using judgments of the form
\begin{center}
$\typeenv{e}{T}$
\end{center}
where
\begin{itemize}
\item $\Gamma$ is the type environment, mapping the names of local variables and global functions
to their types,
\item $\Phi$ is the region access privileges under which the expression will be evaluated, and
\item $\Omega$ is the set of constraints that must hold between regions
\end{itemize}
As regions are hierarchical, privileges and/or constraints on a region often imply similar
privileges or constraints on subregions.  Figure~\ref{fig:closure} defines closure
operations for $\Phi^*$ and $\Omega^*$ that add implied constraints.
The type checking rules are in Figure~\ref{fig:types}.  Many of the rules are 
straightforward, but several deserve more explanation.

Core Legion has no subtyping, even for pointers into regions that are part of a 
known region hierarchy.  The region ``scope'' of a pointer can only be changed via explicit
{\em upregion} and {\em downregion} operations, which check that the ``from'' and ``to'' regions
are in the right relationship.  It would have been possible to safely infer {\em upregion} calls,
but {\em downregion} calls must be accompanied by a run-time check that the pointer really does
point into the subregion(s), so {\em upregion} was made explicit as well for simplicity.

@@

Due to space constraints, we limit the discussion to the general strategy of the proof.  The full
proof can be found here\cite{LegionTypes12}.  If $v$ and $E$ are the result of a permissible evaluation of
$e$, then a proof exists of:
$$\opsenv{e}{v}{E}$$
The soundness of the result is shown by induction over the structure of this proof.  For each of 
the 23 different forms a Legion expression can have, we assume the soundness of the subexpressions'
evaluations (provided we can show the consistency of the mappings and store used by that
subexpression) and prove that the result of the expression's evaluation is sound.  Although there
are a multitude of cases to be proven, many of them are similar, and benefit from
the use of the following lemmas:

@@

The proofs of these lemmas can also be found here\cite{LegionTypes12}.  The proof of the theorem is organized
into six pieces, one for each of the three consistency properties needed for subexpressions and one
for each of the three soundness properties we wish to show.
